true

### 2023-07-10 jss/cheem_paper.tex
failed attempt to go to springer nature template for Computational Statistics.

\documentclass[
]{jss}

%% recommended packages
\usepackage{orcidlink,thumbpdf,lmodern}

@@

\author{
Nicholas Spyrison~\orcidlink{0000-0002-8417-0212}\\Monash
University \And Dianne Cook~\orcidlink{0000-0002-3813-7155}\\Monash
University \And Przemyslaw
Biecek~\orcidlink{0000-0001-8423-1823}\\University of Warsaw\\
Warsaw University of Technology
}
\title{Exploring Local Explanations of Nonlinear Models Using Animated
Linear Projections}

\Plainauthor{Nicholas Spyrison, Dianne Cook, Przemyslaw Biecek}
\Plaintitle{Exploring Local Explanations of Nonlinear Models Using
Animated Linear Projections}
\Shorttitle{Exploring Local Explanations}

\Abstract{
The increased predictive power of machine learning models comes at the
cost of increased complexity and loss of interpretability, particularly
in comparison to parametric statistical models. This trade-off has led
to the emergence of eXplainable AI (XAI) which provides methods, such as
local explanations (LEs) and local variable attributions (LVAs), to shed
light on how a model use predictors to arrive at a prediction. These
provide a point estimate of the linear variable importance in the
vicinity of a single observation. However, LVAs tend not to effectively
handle association between predictors. To understand how the interaction
between predictors affects the variable importance estimate, we can
convert LVAs into linear projections and use the radial tour. This is
also useful for learning how a model has made a mistake, or the effect
of outliers, or the clustering of observations. The approach is
illustrated with examples from categorical (penguin species, chocolate
types) and quantitative (soccer/football salaries, house prices)
response models. The methods are implemented in the R package cheem,
available on CRAN.
}

\Keywords{explainable artificial intelligence, nonlinear model
interpretability, visual analytics, local explanations, grand
tour, radial tour}
\Plainkeywords{explainable artificial intelligence, nonlinear model
interpretability, visual analytics, local explanations, grand
tour, radial tour}

%% publication information
%% \Volume{50}
%% \Issue{9}
%% \Month{June}
%% \Year{2012}
%% \Submitdate{}
%% \Acceptdate{2012-06-04}

\Address{
    Nicholas Spyrison\\
    Monash University\\
    Dept of Human Centred Computing\\
Faculty of Information Technology\\
  E-mail: \email{spyrison@gmail.com}\\
  URL: \url{https://nspyrison.netlify.app}\\~\\
      Dianne Cook\\
    Monash University\\
    Dept of Econometrics and Business Statistics\\

% tightlist command for lists without linebreak
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\usepackage{amsmath} \usepackage{datetime}

\begin{document}

\hypertarget{sec:intro}{%
\section{Introduction}\label{sec:intro}}

There are different reasons and purposes for fitting a model. According
to the taxonomies of \citet{breiman_statistical_2001} and
\citet{shmueli_explain_2010}, it can be useful to group models into two
types: explanatory and predictive. Explanatory modeling is used for
inferential purposes, while predictive modeling focuses solely on the
performance of an objective function. The intended use of the model has
important implications for its selection and development.
Interpretability is critical in explanatory modeling to draw meaningful
inferential conclusions, such as which variables most contribute to a
prediction or whether some observations are less well fit.
Interpretability becomes more difficult when the model is nonlinear.
Nonlinear models occur in statistical models with polynomial or
interaction terms between quantitative predictors, and almost all
computational models such as random forests, support vector machines, or
neural networks
\citep[e.g.][]{breiman_random_2001, boser_training_1992, anderson_introduction_1995}.

In linear models interpretation of the importance of variables is
relatively straightforward, one adjusts for the covariance of multiple
variables when examining the relationship with the response. The
interpretation is valid for the full domain of the predictors. In
nonlinear models, one needs to consider the model in small neighborhoods
of the domain to make any assessment of variable importance. Even though
this is difficult, it is especially important to interpret model fits as
we become more dependent on nonlinear models for routine aspects of life
to avoid issues described in \citet{stahl-ethics}. Understanding how
nonlinear models behave when usage extrapolates outside the domain of
predictors, either in sub-spaces where few samples were provided in the
training set, or extending outside the domain. It is especially
important because nonlinear models can vary wildly and predictions can
be dramatically wrong in these areas.

Explainable Artificial Intelligence (XAI) is an emerging field of
research focused on methods for the interpreting of models
\citep{adadi_peeking_2018, arrieta_explainable_2020}. A class of
techniques, called \emph{local explanations} (LEs), provide methods to
approximate linear variable importance, called local variable
attributions (LVAs), at the location of each observation or the
predictions at a specific point in the data domain. Because these are
point-specific, it is challenging to comprehensively visualize them to
understand a model. There are common approaches for visualizing
high-dimensional data as a whole, but what is needed are new approaches
for viewing these individual LVAs relative to the whole.

For multivariate data visualization, a \emph{tour}
\citep{asimov_grand_1985, buja_grand_1986, lee_state_2021} of linear
data projections onto a lower-dimensional space, could be an element of
XAI, complementing LVAs. Applying tours to model interpretation is
recommended by \citet{wickham_visualizing_2015} primarily to examine the
fitted model in the space of the data. \citet{cook_interactive_2007}
describe the use of tours for exploring classification boundaries and
model diagnostics
\citep{Caragea2008, lee_pptree_2013, da_silva_projection_2021}. There
are various types of tours. In a \emph{manual} or radial tour
\citep{cook_manual_1997, spyrison_spinifex_2020}, the path of linear
projections is defined by changing the contribution of a selected
variable. We propose to use this to scrutinize the LVAs. This approach
could be considered to be a counter-factual, what-if analysis, such as
\emph{ceteris paribus} (``other things held constant'') profiles
\citep{biecek_ceterisparibus_2020}.

@@

LVAs shed light on machine learning model fits by estimating linear
variable importance in the vicinity of a single observation. There are
many approaches for calculating LVAs. A comprehensive summary of the
taxonomy of currently available methods is provided in Figure 6 by
\citet{arrieta_explainable_2020}. It includes a large number of
model-specific explanations such as deepLIFT
\citep{shrikumar_not_2016, shrikumar_learning_2017}, a popular recursive
method for estimating importance in neural networks. There are fewer
model-agnostic methods, of which LIME \citep{ribeiro_why_2016} and
SHaply Additive exPlanations (SHAP) \citep{lundberg_unified_2017}, are
popular.

These observation-level explanations are used in various ways depending
on the data. In image classification, where pixels correspond to
predictors, saliency maps overlay or offset a heatmap to indicate
important pixels \citep{simonyan_deep_2014}. For example, pixels
corresponding to snow may be highlighted as important contributors when
distinguishing if a picture contains a coyote or husky. In text
analysis, word-level contextual sentiment analysis highlights the
sentiment and magnitude of influential words \citep{vanni_textual_2018}.
In the case of numeric regression, they are used to explain additive
contributions of variables from the model intercept to the observation's
prediction \citep{ribeiro_why_2016}.

We will be focusing on SHAP values in this paper, but the approach is
applicable to any method used to calculate the LVAs. SHAP calculates the
variable contributions of one observation by examining the effect of
other variables on the predictions. The term ``SHAP'' refers to
\citet{shapley_value_1953}'s method to evaluate an individual's
contribution in cooperative games by assessing this player's performance
in the presence or absence of other players.
\citet{strumbelj_efficient_2010} introduced SHAP for LEs in machine
learning models. Variable importance can depend on the sequence in which
variables are entered into the model fitting process, thus for any
sequence we get a set of variable contribution values for a single
observation. These values will add up to the difference between the
fitted value for the observation, and the average fitted value for all
observations. Using all possible sequences, or permutations, gives
multiple values for each variable, which are averaged to get the SHAP
value for an observation. It can be helpful to standardize variables
prior to computing SHAP values if they have been measured on different
scales.

The approach is related to partial dependence plots (for example see
chapter 8 of \citet{molnar2022}), used to explain the effect of a
variable by predicting the response for a range of values on this
variable after fixing the value of all other variables to their mean.
Though partial dependence plots are a global approximation of the
variable importance, while SHAP is specific to one observation.

\begin{CodeChunk}
\begin{figure}

@@

\caption[Illustration of SHAP values for a random forest model FIFA 2020 player wages from nine skill predictors]{Illustration of SHAP values for a random forest model FIFA 2020 player wages from nine skill predictors. A star offensive and defensive player are compared, L. Messi and V. van Dijk, respectively. Panel (a) shows breakdown plots of three sequences of the variables. The sequence of the variables impacts the magnitude of their attribution. Panel (b) shows the distribution of attribution for each variable across 25 sequences of predictors, with the mean displayed as a dot for each player. Reaction skills are important for both players. Offense and movement are important for Messi but not van Dijk, and conversely, defense and power are important for van Dijk but not Messi.}\label{fig:shapdistrbd}
\end{figure}
\end{CodeChunk}

We use 2020 season FIFA data \citep{leone_fifa_2020} to illustrate SHAP
following the procedures described in \citet{biecek_explanatory_2021}.
There are 5000 observations of nine predictor variables measuring
players' skills and one response variable, wages (in euros). A random
forest model is fit regressing players' wages on the skill variables. In
this illustration in Figure \ref{fig:shapdistrbd} the SHAP values are
compared for a star offensive player (L. Messi) and a prominent
defensive player (V. van Dijk). We are interested in knowing how the
skill variables locally contribute to the wage prediction of each
player. A difference in the attribution of the variable importance
across the two positions of the players can be expected. This would be
interpreted as how a player's salary depends on which combination of
skills. Panel (a) is a version of a breakdown plot
\citep{gosiewska_ibreakdown_2019} where just three sequences of
variables are shown, for two observations. A breakdown plot shows the
absolute values of the variable attribution for an observation, usually
sorted from the highest value to the lowest. There is no scale on the
horizontal axis here because values are considered relative to each
other. Here we can see how the variable contribution can change
depending on sequence, relative to both players. (Note that the order of
the variables is different in each plot because they have been sorted by
the biggest average contribution across both players.) For all
sequences, and for both players \texttt{reaction} has the strongest
contribution, with perhaps more importance for the defensive player.
Then it differs by player: for Messi \texttt{offense} and
\texttt{movement} have the strongest contributions, and for van Dijk it
is \texttt{defense} and \texttt{power}, regardless of the variable
sequence.

@@

This can also be considered similar to examining the coefficients from
all subsets regression, as described in
\citet{wickham_visualizing_2015}. Various models that are similarly good
might use different combinations of the variables. Examining the
coefficients from multiple models helps to understand the relative
importance of each variable in the context of all other variables. This
is similar to the approach here with SHAP values, that by examining the
variation in values across different permutations of variables, we can
gain more understanding of the relationship between the response and
predictors.

For the application, we use \emph{tree SHAP}, a variant of SHAP that
enjoys a lower computational complexity
\citep{lundberg_consistent_2018}. Instead of aggregating over sequences
of the variables, tree SHAP calculates observation-level variable
importance by exploring the structure of the decision trees. Tree SHAP
is only compatible with tree-based models. so random forests are used
for illustration.

There are numerous R packages currently available that provide functions
for computing SHAP values, including \texttt{fastshap} \citep{fastshap},
\texttt{kernelshap} \citep{kernelshap}, \texttt{shapr} \citep{shapr},
\texttt{shapviz} \citep{shapviz}, \texttt{PPtreeregViz}
\citep{PPtreeregViz}, \texttt{ExplainPrediction}
\citep{ExplainPrediction}, \texttt{flashlight} \citep{flashlight}, and
the package \texttt{DALEX} has many resources \citep{biecek_dalex_2018}.
There are many more packages only available through Github, like
\texttt{treeshap} \citep{kominsarczyk_treeshap_2021} that is used for
this work. \citet{molnar2022} provides good explanations of the
different methods and how to apply them to different models.

@@

A \emph{tour} enables the viewing of high-dimensional data by animating
many linear projections with small incremental changes. It is achieved
by following a path of linear projections (bases) of high-dimensional
space. One key variable of the tour is the object permanence of the data
points; one can track the relative change of observations in time and
gain information about the relationships between points across multiple
variables. There are various types of tours that are distinguished by
how the paths are generated \citep{lee_state_2021, cook_grand_2008}.

The manual tour \citep{cook_manual_1997} defines its path by changing a
selected variable's contribution to a basis to allow the variable to
contribute more or less to the projection. The requirement constrains
the contribution of all other variables that a basis needs to be
orthonormal (columns correspond to vectors, with unit length, and
orthogonal to each other). The manual tour is primarily used to assess
the importance of a variable to the structure visible in a projection.
It also lends itself to pre-computation queued in advance or computed on
the fly for human-in-the-loop analysis
\citep{karwowski_international_2006}.

A version of the manual tour called a \emph{radial tour} is implemented
in \citet{spyrison_spinifex_2020} and forms the basis of this new work.
In a radial tour, the selected variable can change its magnitude of
contribution but not its angle; it must move along the direction of its
original contribution. The implementation allows for pre-computation and
interactive re-calculation to focus on a different variable.

\begin{CodeChunk}
\begin{figure}

@@

\caption{The radial tour allows the user to remove a variable from a projection, to examine the importance of this variable to the structure in the plot. Here we have a 1D projection of the penguins data displayed as a density plot. The line segments on the bottom correspond to the coefficients of the variables making up the projection. The structure in the plot is bimodality (left), and the importance of the variable \textsf{bd} is being explored. As this variable contribution is reduced in the plot (middle, right) we can see that the bimodality decreases. Thus \textsf{bd} is an important variable contributing to the bimodal structure.}\label{fig:radialtour}
\end{figure}
\end{CodeChunk}

@@

To explore the LVAs, coordinated views \citep{roberts_state_2007}
\citep[also known as ensemble graphics,][]{unwin_ensemble_2018} are
provided in the \emph{cheem viewer} application. There are two primary
plots: the \textbf{global view} to give the context of all of the SHAP
values and the \textbf{radial tour view} to explore the LVAs with
user-controlled rotation. There are numerous user inputs, including
variable selection for the radial tour and observation selection for
making comparisons. There are different plots used for the categorical
and quantitative responses. Figures \ref{fig:classificationcase} and
\ref{fig:regressioncase} are screenshots showing the cheem viewer for
the two primary tasks: classification (categorical response) and
regression (quantitative response).

@@

The LVAs for all observations are normalized (sum of squares equals 1),
and thus, the relative importance of variables can be compared across
all observations. These are depicted as a vertical parallel coordinate
plot \citep{ocagne_coordonnees_1885}. (The SHAP values of the PI and CI
are shown as dashed and dotted lines, respectively.) One should obtain a
sense of the overall importance of variables from this plot. The more
important variables will have larger values, and in the case of
classification tasks variables that have different magnitudes for
different classes are more globally important. For example, Figure
\ref{fig:classificationcase}e suggests that \texttt{bl} is important for
distinguishing the green class from the other two. For regression, one
might generally observe which variables have low values for all
observations (not important). For example, \texttt{BMI} and \texttt{pwr}
in Figure \ref{fig:regressioncase}e, have a range of high and low values
(e.g., \texttt{off}, \texttt{def}), suggesting they are important for
some observations and not important for others.

@@

\begin{CodeChunk}
\begin{figure}

@@

\renewcommand\refname{References}
\bibliography{paper.bib}

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}

@@

  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdfkeywords={true},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{true}
\author{true \and true \and true}
\date{}

\begin{document}
\maketitle
\begin{abstract}
The increased predictive power of machine learning models comes at the cost of increased complexity and loss of interpretability, particularly in comparison to parametric statistical models. This trade-off has led to the emergence of eXplainable AI (XAI) which provides methods, such as local explanations (LEs) and local variable attributions (LVAs), to shed light on how a model use predictors to arrive at a prediction. These provide a point estimate of the linear variable importance in the vicinity of a single observation. However, LVAs tend not to effectively handle association between predictors. To understand how the interaction between predictors affects the variable importance estimate, we can convert LVAs into linear projections and use the radial tour. This is also useful for learning how a model has made a mistake, or the effect of outliers, or the clustering of observations. The approach is illustrated with examples from categorical (penguin species, chocolate types) and quantitative (soccer/football salaries, house prices) response models. The methods are implemented in the R package cheem, available on CRAN.
\end{abstract}

{
\setcounter{tocdepth}{2}
\tableofcontents
}
\hypertarget{sec:intro}{%
\section{Introduction}\label{sec:intro}}

There are different reasons and purposes for fitting a model. According to the taxonomies of Breiman (2001b) and Shmueli (2010), it can be useful to group models into two types: explanatory and predictive. Explanatory modeling is used for inferential purposes, while predictive modeling focuses solely on the performance of an objective function. The intended use of the model has important implications for its selection and development. Interpretability is critical in explanatory modeling to draw meaningful inferential conclusions, such as which variables most contribute to a prediction or whether some observations are less well fit. Interpretability becomes more difficult when the model is nonlinear. Nonlinear models occur in statistical models with polynomial or interaction terms between quantitative predictors, and almost all computational models such as random forests, support vector machines, or neural networks (e.g. Breiman 2001a; Boser, Guyon, and Vapnik 1992; Anderson 1995).

In linear models interpretation of the importance of variables is relatively straightforward, one adjusts for the covariance of multiple variables when examining the relationship with the response. The interpretation is valid for the full domain of the predictors. In nonlinear models, one needs to consider the model in small neighborhoods of the domain to make any assessment of variable importance. Even though this is difficult, it is especially important to interpret model fits as we become more dependent on nonlinear models for routine aspects of life to avoid issues described in Stahl (2021). Understanding how nonlinear models behave when usage extrapolates outside the domain of predictors, either in sub-spaces where few samples were provided in the training set, or extending outside the domain. It is especially important because nonlinear models can vary wildly and predictions can be dramatically wrong in these areas.

Explainable Artificial Intelligence (XAI) is an emerging field of research focused on methods for the interpreting of models (Adadi and Berrada 2018; Barredo Arrieta et al. 2020). A class of techniques, called \emph{local explanations} (LEs), provide methods to approximate linear variable importance, called local variable attributions (LVAs), at the location of each observation or the predictions at a specific point in the data domain. Because these are point-specific, it is challenging to comprehensively visualize them to understand a model. There are common approaches for visualizing high-dimensional data as a whole, but what is needed are new approaches for viewing these individual LVAs relative to the whole.

For multivariate data visualization, a \emph{tour} (Asimov 1985; Buja and Asimov 1986; S. Lee et al. 2021) of linear data projections onto a lower-dimensional space, could be an element of XAI, complementing LVAs.
Applying tours to model interpretation is recommended by Wickham, Cook, and Hofmann (2015) primarily to examine the fitted model in the space of the data. Cook, Swayne, and Buja (2007) describe the use of tours for exploring classification boundaries and model diagnostics (Caragea et al. 2008; Y. D. Lee et al. 2013; da Silva, Cook, and Lee 2021).
There are various types of tours. In a \emph{manual} or radial tour (Cook and Buja 1997; Spyrison and Cook 2020), the path of linear projections is defined by changing the contribution of a selected variable. We propose to use this to scrutinize the LVAs.
This approach could be considered to be a counter-factual, what-if analysis, such as \emph{ceteris paribus} (``other things held constant'') profiles (Biecek 2020).

@@

LVAs shed light on machine learning model fits by estimating linear variable importance in the vicinity of a single observation. There are many approaches for calculating LVAs.
A comprehensive summary of the taxonomy of currently available methods is provided in Figure 6 by Barredo Arrieta et al. (2020). It includes a large number of model-specific explanations such as deepLIFT (Shrikumar et al. 2016; Shrikumar, Greenside, and Kundaje 2017), a popular recursive method for estimating importance in neural networks. There are fewer model-agnostic methods, of which LIME (Ribeiro, Singh, and Guestrin 2016) and SHaply Additive exPlanations (SHAP) (Lundberg and Lee 2017), are popular.

These observation-level explanations are used in various ways depending on the data. In image classification, where pixels correspond to predictors, saliency maps overlay or offset a heatmap to indicate important pixels (Simonyan, Vedaldi, and Zisserman 2014). For example, pixels corresponding to snow may be highlighted as important contributors when distinguishing if a picture contains a coyote or husky. In text analysis, word-level contextual sentiment analysis highlights the sentiment and magnitude of influential words (Vanni et al. 2018). In the case of numeric regression, they are used to explain additive contributions of variables from the model intercept to the observation's prediction (Ribeiro, Singh, and Guestrin 2016).

We will be focusing on SHAP values in this paper, but the approach is applicable to any method used to calculate the LVAs. SHAP calculates the variable contributions of one observation by examining the effect of other variables on the predictions. The term ``SHAP'' refers to Shapley (1953)'s method to evaluate an individual's contribution in cooperative games by assessing this player's performance in the presence or absence of other players. Strumbelj and Kononenko (2010) introduced SHAP for LEs in machine learning models. Variable importance can depend on the sequence in which variables are entered into the model fitting process, thus for any sequence we get a set of variable contribution values for a single observation. These values will add up to the difference between the fitted value for the observation, and the average fitted value for all observations. Using all possible sequences, or permutations, gives multiple values for each variable, which are averaged to get the SHAP value for an observation. It can be helpful to standardize variables prior to computing SHAP values if they have been measured on different scales.

The approach is related to partial dependence plots (for example see chapter 8 of Molnar (2022)), used to explain the effect of a variable by predicting the response for a range of values on this variable after fixing the value of all other variables to their mean. Though partial dependence plots are a global approximation of the variable importance, while SHAP is specific to one observation.

\begin{figure}

@@

\caption{Illustration of SHAP values for a random forest model FIFA 2020 player wages from nine skill predictors. A star offensive and defensive player are compared, L. Messi and V. van Dijk, respectively. Panel (a) shows breakdown plots of three sequences of the variables. The sequence of the variables impacts the magnitude of their attribution. Panel (b) shows the distribution of attribution for each variable across 25 sequences of predictors, with the mean displayed as a dot for each player. Reaction skills are important for both players. Offense and movement are important for Messi but not van Dijk, and conversely, defense and power are important for van Dijk but not Messi.}\label{fig:shapdistrbd}
\end{figure}

We use 2020 season FIFA data (Leone 2020) to illustrate SHAP following the procedures described in Biecek and Burzykowski (2021). There are 5000 observations of nine predictor variables measuring players' skills and one response variable, wages (in euros). A random forest model is fit regressing players' wages on the skill variables. In this illustration in Figure \ref{fig:shapdistrbd} the SHAP values are compared for a star offensive player (L. Messi) and a prominent defensive player (V. van Dijk). We are interested in knowing how the skill variables locally contribute to the wage prediction of each player. A difference in the attribution of the variable importance across the two positions of the players can be expected. This would be interpreted as how a player's salary depends on which combination of skills. Panel (a) is a version of a breakdown plot (Gosiewska and Biecek 2019) where just three sequences of variables are shown, for two observations. A breakdown plot shows the absolute values of the variable attribution for an observation, usually sorted from the highest value to the lowest. There is no scale on the horizontal axis here because values are considered relative to each other. Here we can see how the variable contribution can change depending on sequence, relative to both players. (Note that the order of the variables is different in each plot because they have been sorted by the biggest average contribution across both players.) For all sequences, and for both players \texttt{reaction} has the strongest contribution, with perhaps more importance for the defensive player. Then it differs by player: for Messi \texttt{offense} and \texttt{movement} have the strongest contributions, and for van Dijk it is \texttt{defense} and \texttt{power}, regardless of the variable sequence.

@@

This can also be considered similar to examining the coefficients from all subsets regression, as described in Wickham, Cook, and Hofmann (2015). Various models that are similarly good might use different combinations of the variables. Examining the coefficients from multiple models helps to understand the relative importance of each variable in the context of all other variables. This is similar to the approach here with SHAP values, that by examining the variation in values across different permutations of variables, we can gain more understanding of the relationship between the response and predictors.

For the application, we use \emph{tree SHAP}, a variant of SHAP that enjoys a lower computational complexity (Lundberg, Erion, and Lee 2018). Instead of aggregating over sequences of the variables, tree SHAP calculates observation-level variable importance by exploring the structure of the decision trees. Tree SHAP is only compatible with tree-based models. so random forests are used for illustration.

There are numerous R packages currently available that provide functions for computing SHAP values, including \texttt{fastshap} (Greenwell 2021), \texttt{kernelshap} (Mayer and Watson 2023), \texttt{shapr} (Sellereite, Jullum, and Redelmeier 2023), \texttt{shapviz} (Mayer 2023b), \texttt{PPtreeregViz} (E.-K. Lee and Cho 2022), \texttt{ExplainPrediction} (Robnik-Sikonja 2018), \texttt{flashlight} (Mayer 2023a), and the package \texttt{DALEX} has many resources (Biecek 2018). There are many more packages only available through Github, like \texttt{treeshap} (Kominsarczyk et al. 2021) that is used for this work. Molnar (2022) provides good explanations of the different methods and how to apply them to different models.

@@

A \emph{tour} enables the viewing of high-dimensional data by animating many linear projections with small incremental changes. It is achieved by following a path of linear projections (bases) of high-dimensional space. One key variable of the tour is the object permanence of the data points; one can track the relative change of observations in time and gain information about the relationships between points across multiple variables. There are various types of tours that are distinguished by how the paths are generated (S. Lee et al. 2021; Cook et al. 2008).

The manual tour (Cook and Buja 1997) defines its path by changing a selected variable's contribution to a basis to allow the variable to contribute more or less to the projection. The requirement constrains the contribution of all other variables that a basis needs to be orthonormal (columns correspond to vectors, with unit length, and orthogonal to each other). The manual tour is primarily used to assess the importance of a variable to the structure visible in a projection. It also lends itself to pre-computation queued in advance or computed on the fly for human-in-the-loop analysis (Karwowski 2006).

A version of the manual tour called a \emph{radial tour} is implemented in Spyrison and Cook (2020) and forms the basis of this new work. In a radial tour, the selected variable can change its magnitude of contribution but not its angle; it must move along the direction of its original contribution. The implementation allows for pre-computation and interactive re-calculation to focus on a different variable.

\begin{figure}

@@

\caption{The radial tour allows the user to remove a variable from a projection, to examine the importance of this variable to the structure in the plot. Here we have a 1D projection of the penguins data displayed as a density plot. The line segments on the bottom correspond to the coefficients of the variables making up the projection. The structure in the plot is bimodality (left), and the importance of the variable \textsf{bd} is being explored. As this variable contribution is reduced in the plot (middle, right) we can see that the bimodality decreases. Thus \textsf{bd} is an important variable contributing to the bimodal structure.}\label{fig:radialtour}
\end{figure}

@@

To explore the LVAs, coordinated views (Roberts 2007) (also known as ensemble graphics, Unwin and Valero-Mora 2018) are provided in the \emph{cheem viewer} application. There are two primary plots: the \textbf{global view} to give the context of all of the SHAP values and the \textbf{radial tour view} to explore the LVAs with user-controlled rotation. There are numerous user inputs, including variable selection for the radial tour and observation selection for making comparisons. There are different plots used for the categorical and quantitative responses. Figures \ref{fig:classificationcase} and \ref{fig:regressioncase} are screenshots showing the cheem viewer for the two primary tasks: classification (categorical response) and regression (quantitative response).

@@

The LVAs for all observations are normalized (sum of squares equals 1), and thus, the relative importance of variables can be compared across all observations. These are depicted as a vertical parallel coordinate plot (Ocagne 1885). (The SHAP values of the PI and CI are shown as dashed and dotted lines, respectively.) One should obtain a sense of the overall importance of variables from this plot. The more important variables will have larger values, and in the case of classification tasks variables that have different magnitudes for different classes are more globally important. For example, Figure \ref{fig:classificationcase}e suggests that \texttt{bl} is important for distinguishing the green class from the other two. For regression, one might generally observe which variables have low values for all observations (not important). For example, \texttt{BMI} and \texttt{pwr} in Figure \ref{fig:regressioncase}e, have a range of high and low values (e.g., \texttt{off}, \texttt{def}), suggesting they are important for some observations and not important for others.

@@

\begin{figure}

@@

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}
